---
title: "Analyzing trends of organ donor enrollment in New York state"
output: 
  github_document:
    md_extensions: +tex_math_dollars
                   +raw_html
always_allow_html: yes
---

```{r setup and data import, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, dpi = 300, out.width = "75%")

library(tidyverse)
library(lme4)
library(modelr)
library(kableExtra)
#using data cleaned in api_pull.rmd file

theme_set(theme_bw())

organ_sp <- read_csv("data/organ_spline.csv") %>% 
  filter(!(county %in% c("TOTAL NYS", "Cattauragus", "St Lawrence"))) %>% 
  mutate(opo = as_factor(opo), 
         opo = fct_relevel(opo, 
                           "New York Organ Donor Network", 
                           "Center for Donation and Transplant in New York", 
                           "UNYTS", 
                           "Finger Lakes Donor Recovery Network"))
```

|Name            | Uni       |
|----------------|-----------|
|Shuliang Deng   | sd3258    |
|Nick Williams   | ntw2117   |
|Sijia Yue       | sy2824    |
|Jack Yan        | xy2395    |
|Alina Levine    | al3851    |

[Click here for data](https://www.dropbox.com/sh/grlugs9hrksvkqr/AAAwILSg5UyMrW1CD7dWmHPNa?dl=0)

## Motivation

## Initial questions

## Data 

## Exploratory analyses

## Formal analyses

Formal analyses of the effect of major policy changes were conducted using mixed-effects linear models with a random intercept; counties were treated as the clustering variable. The outcome was analyzed was the proportion of eligible adults registered with one of four organ donor organizations. Four successive models were created: a model only analyzing a linear effect of time; a model including a term for a spline corresponding with a 2012 policy INSERT WHAT THIS POLICY WAS; a model including a term for a spline corresponding with a 2016 policy INSERT WHAT THIS POLICY WAS; a model including a term for a spline corresponding with a 2017 policy that lowered the age from eligible enrollees from 18 to 16 years old; and a model that included all splines. All models were adjusted for the specific organization used in each county; models were compared using likelihood ratio tests. 

```{r overall time, echo = FALSE}
no_sp_model <- organ_sp %>% 
  lmer(eligible_population_enrolled ~ total_days + opo + (1 | county), data = ., 
       REML = FALSE)

no_sp_model %>% 
  broom::tidy() %>% 
  filter(term == "total_days") %>% 
  mutate(estimate_year = estimate * 365, 
         se_year = std.error * 365, 
         ci_low = estimate_year - 1.96*se_year, 
         ci_upper = estimate_year + 1.96*se_year) %>% 
  mutate(term = ifelse(term == "total_days", "Time (year)", term)) %>% 
  select(term, estimate_year, se_year, ci_low, ci_upper) %>% 
  rename("Term" = term, 
         "Coefficient" = estimate_year, 
         "SE" = se_year, 
         "Lower bound" = ci_low, 
         "Upper bound" = ci_upper) %>% 
  knitr::kable(digits = 3, escape = F) %>% 
  kable_styling(full_width = F, bootstrap_options = c("hover", "striped"))

add_predictions(organ_sp, no_sp_model) %>% 
  ggplot(aes(x = total_days, y = pred, color = opo, type = county)) + 
  geom_line() + 
  viridis::scale_color_viridis(discrete = TRUE) + 
  theme(legend.position = "bottom") + 
  guides(color = guide_legend(ncol = 2))
```

```{r spline 2012, echo = FALSE}
sp_12_model <- organ_sp %>% 
  lmer(eligible_population_enrolled ~ total_days + year_sp_2012 + (1 | county), data = .)

sp_12_model %>% 
  broom::tidy() %>% 
  filter(term %in% c("total_days", "year_sp_2012")) %>% 
  mutate(estimate_year = estimate * 365, 
         se_year = std.error * 365, 
         ci_low = estimate_year - 1.96*se_year, 
         ci_upper = estimate_year + 1.96*se_year) %>% 
  mutate(term = ifelse(term == "total_days", "Time (year)", term), 
         term = ifelse(term == "year_sp_2012", "2012 spline", term)) %>% 
  select(term, estimate_year, se_year, ci_low, ci_upper) %>% 
  rename("Term" = term, 
         "Coefficient" = estimate_year, 
         "SE" = se_year, 
         "Lower bound" = ci_low, 
         "Upper bound" = ci_upper) %>% 
  knitr::kable(digits = 3) %>% 
  kable_styling(full_width = F, bootstrap_options = c("hover", "striped"))

add_predictions(organ_sp, sp_12_model) %>% 
  ggplot(aes(x = total_days, y = pred, color = opo, type = county)) + 
  geom_line() +
  viridis::scale_color_viridis(discrete = TRUE) + 
  theme(legend.position = "bottom")+ 
  guides(color = guide_legend(ncol = 2))
```

```{r spline 2016, echo = FALSE}
sp_16_model <- organ_sp %>% 
  lmer(eligible_population_enrolled ~ total_days + year_sp_2016 + opo + (1 | county), data = .) 

sp_16_model %>% 
  broom::tidy() %>% 
  filter(term %in% c("total_days", "year_sp_2016")) %>% 
  mutate(estimate_year = estimate * 365, 
         se_year = std.error * 365, 
         ci_low = estimate_year - 1.96*se_year, 
         ci_upper = estimate_year + 1.96*se_year) %>% 
  mutate(term = ifelse(term == "total_days", "Time (year)", term), 
         term = ifelse(term == "year_sp_2016", "2016 spline", term)) %>% 
  select(term, estimate_year, se_year, ci_low, ci_upper) %>% 
  rename("Term" = term, 
         "Coefficient" = estimate_year, 
         "SE" = se_year, 
         "Lower bound" = ci_low, 
         "Upper bound" = ci_upper) %>% 
  knitr::kable(digits = 3) %>% 
  kable_styling(full_width = F, bootstrap_options = c("hover", "striped"))

add_predictions(organ_sp, sp_16_model) %>% 
  ggplot(aes(x = total_days, y = pred, color = opo, type = county)) + 
  geom_line() + 
  viridis::scale_color_viridis(discrete = TRUE) + 
  theme(legend.position = "bottom")+ 
  guides(color = guide_legend(ncol = 2))
```

```{r spline 2017, echo = FALSE}
sp_17_model <- organ_sp %>% 
  lmer(eligible_population_enrolled ~ total_days + year_sp_2017 + opo + (1 | county), data = ., 
       REML = FALSE)

sp_17_model %>% 
  broom::tidy() %>% 
  filter(term %in% c("total_days", "year_sp_2017")) %>% 
  mutate(estimate_year = estimate * 365, 
         se_year = std.error * 365, 
         ci_low = estimate_year - 1.96*se_year, 
         ci_upper = estimate_year + 1.96*se_year) %>% 
  mutate(term = ifelse(term == "total_days", "Time (year)", term), 
         term = ifelse(term == "year_sp_2017", "2017 spline", term)) %>% 
  select(term, estimate_year, se_year, ci_low, ci_upper) %>% 
  rename("Term" = term, 
         "Coefficient" = estimate_year, 
         "SE" = se_year, 
         "Lower bound" = ci_low, 
         "Upper bound" = ci_upper) %>% 
  knitr::kable(digits = 3) %>% 
  kable_styling(full_width = F, bootstrap_options = c("hover", "striped"))

add_predictions(organ_sp, sp_17_model) %>% 
  ggplot(aes(x = total_days, y = pred, color = opo, type = county)) + 
  geom_line() +
  viridis::scale_color_viridis(discrete = TRUE) + 
  theme(legend.position = "bottom") + 
  guides(color = guide_legend(ncol = 2))
```

```{r all splines, echo = FALSE}
all_sp_model <- organ_sp %>% 
  lmer(eligible_population_enrolled ~ total_days + year_sp_2012 + year_sp_2016 + year_sp_2017 + opo + 
         (1 | county), data = ., REML = FALSE)

all_sp_model %>% 
  broom::tidy() %>% 
  filter(term %in% c("total_days", "year_sp_2012", "year_sp_2016", "year_sp_2017")) %>% 
  mutate(estimate_year = estimate * 365, 
         se_year = std.error * 365, 
         ci_low = estimate_year - 1.96*se_year, 
         ci_upper = estimate_year + 1.96*se_year) %>% 
  mutate(term = ifelse(term == "total_days", "Time (year)", term), 
         term = ifelse(term == "year_sp_2012", "2012 spline", term), 
         term = ifelse(term == "year_sp_2016", "2016 spline", term), 
         term = ifelse(term == "year_sp_2017", "2017 spline", term)) %>% 
  select(term, estimate_year, se_year, ci_low, ci_upper) %>% 
  rename("Term" = term, 
         "Coefficient" = estimate_year, 
         "SE" = se_year, 
         "Lower bound" = ci_low, 
         "Upper bound" = ci_upper) %>% 
  knitr::kable(digits = 3) %>% 
  kable_styling(full_width = F, bootstrap_options = c("hover", "striped")) %>% 
  footnote(general = "Splines correspond with dates in which 3 major policy changes were inacted. \nEstimates are adjusted for organ donor organization")

add_predictions(organ_sp, all_sp_model) %>% 
  ggplot(aes(x = total_days, y = pred, color = opo, type = county)) + 
  geom_line() +
  viridis::scale_color_viridis(discrete = TRUE) + 
  geom_vline(xintercept = 1491, linetype = "dashed") + 
  geom_vline(xintercept = 2799, linetype = "dashed") + 
  geom_vline(xintercept = 3088, linetype = "dashed") + 
  annotate(geom = "text", x = 1720, y = 5, label = "2012") +
  annotate(geom = "text", x = 2600, y = 5, label = "2016") +
  annotate(geom = "text", x = 3300, y = 9, label = "2017") +
  theme(legend.position = "bottom", 
        legend.title = element_blank()) + 
  guides(color = guide_legend(ncol = 2)) + 
  labs(title = "Proportion of organ donors in NY counties", 
       x = "Days since start of data collection",
       y = "% enrolled as donors")
```


## Discussion
